Summary of the things that I learned

### 2024-12-09 600-paper.tex
Improve notation and wording

Now, Scott's representation theorem, which asks whether we can represent any $ \lambda $-theory as the endomorphism theory of some reflexive object in some cartesian closed category, can be answered using the presheaf category.
\begin{theorem}[\coqident{AlgebraicTheories.RepresentationTheorem}{representation_theorem_iso}]\label{thm:representation-theorem}
  Any $ \lambda $-theory $ L $ is isomorphic to the endomorphism $ \lambda $-theory $ E_{\Pshf L}(L) $, with $ L $ viewed as a presheaf.
\end{theorem}
\begin{proof}
  First of all, remember that $ L $ is indeed exponentiable and that $ L^L = A(L, 1) $.
  Now, since $ L $ is a $ \lambda $-theory, we have sequences of functions back and forth $ \lambda_n: A(L, 1)_n \to L_n $ and $ \rho_n: L_n \to A(L, 1)_n $. These commute with the $ L $-actions, so they constitute presheaf morphisms and $ E(L) $ is indeed a $ \lambda $-theory.

  Remark \ref{rem:presheaf-Yoneda} gives a sequence of bijections $ \varphi_n : \Pshf L(L^n, L) \cong L_n $ for all $ n $, sending $ F: \Pshf L(L^n, L) $ to $ F(x_1, \dots, x_n) $, and conversely sending $ s: L_n $ to $ ((t_1, \dots, t_n) \mapsto s \bullet (t_1, \dots, t_n)) $. It considers $ \lambda $-terms in $ n $ variables as $ n $-ary functions on the $ \lambda $-calculus. Therefore, it should come as no surprise that $ \varphi $ preserves the $ x_i $, $ \bullet $, $ \rho $ and $ \lambda $, which makes it into an isomorphism of $ \lambda $-theories and this concludes the proof.
\end{proof}
So Hyland shows that the representation theorem follows largely from the fact that the functions from $ L_n $ to itself can be represented by $ L_{n + 1} $, together with the Yoneda lemma for the Lawvere theory associated to $ L $ (Remark \ref{rem:presheaf-Yoneda}).

@@

\begin{remark}
  Actually, in this thesis we stumbled upon another category with a reflexive object that exhibits $ L $ as its endomorphism theory, provided that $ L $ has $ \beta $-equality: In Appendix \ref{ch:weak-cartesian-closed-categories}, we see that the Lawvere theory $ C $ associated to $ L $ is a weak cartesian closed category, in which $ 1^1 = 1 $. It is then easy to see that the endomorphism theory that we obtain from this reflexive object $ 1 $ is exactly equal to $ L $.

  Note, however, that this is technically not an answer to Scott's original theorem, because the original formulation asks for a cartesian closed category with a reflexive object, whereas here we only have a reflexive object in a weak cartesian closed category. Even so, it is an answer to the core of the question, which is about whether we can represent any $ \lambda $-theory as the sets of morphisms $ C(U^n, U) $ for some $ U : C $.
\end{remark}

@@

Note that if we consider $ L_1 $ as a monoid, with operation $ \bullet $ and unit $ x_1 $, and if we consider this monoid as a category, we can embed this (fully faithfully) into $ \mathbf L $. This gives the following sequence of embeddings:
\begin{center}
  \begin{tikzcd}
    L_1 \arrow[r, hookrightarrow] & \mathbf L \arrow[r, hookrightarrow] & \R \arrow[r, hookrightarrow] & \Pshf L
  \end{tikzcd}
\end{center}

Note that the composition of the first two morphisms sends the object of the category $ L_1 $ to $ (\lambda x_1, x_1) : \R $, which is exactly the usual embedding of $ L_1 $ into its Karoubi envelope.

Write $ i: \op{L_1} \hookrightarrow \op{\mathbf L} $ and $ j: \op{\mathbf L} \hookrightarrow \op \R $ for the embeddings on the opposite categories. By Corollary \ref{cor:surjective-precomposition}, the first two functors in this sequence essentially yield surjective functors on the presheaf categories:
\begin{center}
  \begin{tikzcd}
    C_{L_1} \arrow[d, "\yo", hook] &
    \mathbf L \arrow[l, "\op i"'] \arrow[d, "yo", hook] &
    \mathbf R \arrow[rd, hook] \arrow[d, "\yo", hook] \arrow[l, "\op j"']\\
    P C_{L_1} &
    P \mathbf L \arrow[l, "i_*"', two heads] &
    P \R \arrow[l, "j_*"', two heads] \arrow[ll, "\sim"', bend left] &
    \Pshf L \arrow[ll, "\sim", bend left]
  \end{tikzcd}
\end{center}
The two equivalences in this diagram are due to Lemma \ref{lem:lawvere-presheaf} and Corollary \ref{cor:karoubi-presheaf}.

@@

Recall that in the last chapter, we saw that Paul Taylor shows that the category of retracts $ \R $, studied by Scott, is not only cartesian closed, but also `relatively cartesian closed'. To make his proof work in univalent foundations, we need the axiom of choice. The category in question is the Karoubi envelope of the category $ C_{L_0} $ associated to the monoid $ (L_0, \circ) $.

Now, recall from Chapter \ref{ch:category-theory}, that in univalent foundations there are three different definitions that we can take for the Karoubi envelope, denoted $ \overline C_{L_0} $, $ \tilde C_{L_0} $ and $ \hat C_{L_0} $. There is three of them because in classical category theory there are two definitions that are equivalent, and because for one of the definitions we have to choose whether we interpret the `existence' of a retraction as additional structure (the retraction is given explicitly) or as a property (where we demand `mere existence'). This gives the following diagram (Corollary \ref{cor:karoubi-candidates}):
\begin{center}
  \begin{tikzcd}
    C \arrow[r, hook, "\iota_C"] & \overline C \arrow[r, "\sim"] & \tilde C \arrow[r, hook, two heads] & \hat C \arrow[r, phantom, "\subseteq"] & PC
  \end{tikzcd}
\end{center}
where $ \overline C_{L_0} $ and $ \tilde C_{L_0} $ are equivalent to each other, but usually not to $ \hat C_{L_0} $. We saw that $ \hat C_{L_0} $ is the Rezk completion of $ \overline C_{L_0} $ and $ \tilde C_{L_0} $.

@@

The proofs of Scott and Taylor are about $ \overline C_{L_0} $, because they work explicitly with idempotents. The question whether it is possible to lift the proofs to $ \tilde C_{L_0} $ or $ \hat C_{L_0} $, is left for future research. On the other hand, as we will see, Hyland works with presheaves and in this way, he shows that $ \hat C_{L_0} $ is relatively cartesian closed.

Recall that in this thesis, we use the notation $ \Delta_X $ for the functor $ C \to (C \downarrow X) $, sending $ Y $ to $ p_1 : X \times Y \to X $. Also, we use the notation $ X \triangleleft Y $ to denote the type of retraction-section pairs $ r: Y \to X $ and $ s: X \to Y $. We will sometimes not name $ (r, s) : X \triangleleft Y $ explicitly, but just mention that we have $ X \triangleleft Y $, meaning that we have such $ r $ and $ s $.

Now, officially, $ \hat C_{L_0} $ is the full subcategory of $ P C_{L_0} $, consisting of all objects $ X : P C_{L_0} $ such that there merely exists $ X \triangleleft \yo(\star) $. However, recall from the previous section that $ P C_{L_0} \simeq P C_{L_1} \simeq \Pshf L $. Under this equivalence, $ \yo(\star) $ is embedded as the theory presheaf $ L $, and we will work in the full subcategory $ \hat \R \subseteq \Pshf L $ of the objects $ X : \Pshf L $ such that there merely exists $ X \triangleleft L $. Of course, then $ \hat \R $ is equivalent to $ \hat C_{L_0} $.

@@

  Also, functors preserve retractions, so if we have a functor $ F : C \to D $ and we have $ (r, s) : X \triangleleft Y $, then we have
  \[ (F(r), F(s)) : F(X) \triangleleft F(Y). \]

  Lastly, if we want to show that we can do some construction on some $ X : \hat \R $, we can often borrow the construction from $ \Pshf L $ to get some object $ Y : \Pshf L $. For this part, we cannot use the $ (r, s) : X \triangleleft L $. To show that $ Y $ is indeed in $ \hat \R $, we must show that there merely exist $ (r^\prime, s^\prime) : Y \triangleleft L $. By the recursion principle of the propositional truncation, we can assume that we have a concrete $ (r, s) : X \triangleleft L $ to construct $ r^\prime $ and $ s^\prime $.
\end{remark}

Recall that Paul Taylor's display maps for $ \R $ can be characterized as the retracts of $ \Delta_X U $ (the projection $ p_1: X \times U \to X $) in $ (\R \downarrow X) $. This idea also works for $ \hat \R $, and Hyland defines $ D(Y, X) \subseteq \hat \R(Y, X) $ to consist of the retracts of $ \Delta_X L $ in $ (\hat \R \downarrow X) $. Note that, because $ \hat \R $ is a full subcategory of $ \Pshf L $, $ (\hat \R \downarrow X) $ is equivalent to $ (\Pshf L \downarrow X) $ for $ X : \hat \R $.

@@

  Therefore, if we want $ D(Y, X) $ to be a subset of $ \hat \R(Y, X) $, we need to take the propositional truncation of the existence of $ r $ and $ s $.
\end{remark}

\begin{remark}
  Recall that for all $ X : \R $, we have $ X \triangleleft U = \lambda x_1, x_1 $, so by the above, for all $ Y $ in the image of the embedding of $ \R $ into $ \Pshf L $, we have $ Y \triangleleft L $. In particular, $ Y $ is in $ \hat \R $. Actually, we already knew this, because this essentially describes the embedding of $ \R $ into its Rezk completion $ \hat \R $.

  An important example is $ U \times U : \R $, which maps to something isomorphic to $ L \times L $ (because the Yoneda embedding and any adjoint equivalence preserves limits). The retraction $ (r, s) : L \times L \triangleleft L $ is given explicitly by the (pairing and splitting) morphisms
  \[ s_n(a, b) = \lambda x_{n + 1}, x_{n + 1} a b \quad \text{and} \quad r_n(a) = (a (\lambda x_{n + 1} x_{n + 2}, x_{n + 1}), a (\lambda x_{n + 1} x_{n + 2}, x_{n + 2})). \]
  Now, for any $ X, Y : \Pshf L $ and $ f : \Pshf L(Y, X) $, if we have $ (Y, f) \triangleleft \Delta_X L $ in $ \Pshf L $, we have in particular $ Y \triangleleft X \times L $. If we also have $ X \triangleleft L $, we also have $ X \times L \triangleleft L \times L $ because $ - \times L $ is a functor. Then we have
  \[ Y \triangleleft X \times L \triangleleft L \times L \triangleleft L. \]
  In particular, for any $ X : \hat \R $, if some $ (Y, f) : (\Pshf L \downarrow X) $ is a retract of $ \Delta_X L $, we know that $ Y $ is in $ \hat \R $.
\end{remark}

Note that $ D $ is indeed a class of display maps:
\begin{enumerate}
  \item Take $ f : D(Y, X) $ and $ g: \hat \R(Z, X) $. We can borrow the pullback from $ \Pshf L $ to get $ g^*(Y, f) : (\Pshf L \downarrow Z) $. We now must show that there merely exists $ g^* (Y, f) \triangleleft \Delta_Z L $ and in this part, we can assume that the mere existences of retractions are actually given by concrete retraction-section pairs.

    Now, note that taking pullbacks gives a functor $ g^* : (\Pshf L \downarrow X) \to (\Pshf L \downarrow Z) $. Now, $ (Y, f) \triangleleft \Delta_X L $, gives $ (g^*Y, g^*f) \triangleleft g^*(\Delta_X L) $:
    \begin{center}
      \begin{tikzcd}
        g^* Y
          \arrow[rd, "g^*s", shift left]
          \arrow[dd, "g^* f"]
          \arrow[rr] &&
        Y
          \arrow[rd, "s", shift left]
          \arrow[dd, "f"] & \\
        & g^* (X \times L)
          \arrow[lu, "g^*r", shift left]
          \arrow[ld, "g^* p_1"]
          \arrow[rr] & &
        X \times L
          \arrow[ld, "p_1"]
          \arrow[lu, "r", shift left] \\
        Z
          \arrow[rr, "g"] & &
        X &
      \end{tikzcd}
    \end{center}
    The isomorphism between $ g^*(\Delta_X L) $ and $ \Delta_Z L $ (Remark \ref{rem:pullback-of-projection}) shows that we have $ (g^*Y, g^*f) \triangleleft \Delta_Z L $. From this, it also follows that $ g^* Y $ is in $ \hat \R $.

    Now, $ g^*Y $ is the pullback of $ f $ and $ g $ in $ \Pshf L $, and $ \hat \R $ is a full subcategory of $ \Pshf L $, so $ g^* Y $ is also the pullback of $ f $ and $ g $ in $ \hat \R $.

@@

  \item Note that $ \Pshf L $ has a terminal presheaf $ I_n = \{ \star \} $. Under the embedding of $ \R $ into $ \Pshf L $, this is (isomorphic to) the image of $ \lambda x_1 x_2, x_2 : \R $, and therefore, it is in $ \hat \R $. Now, given $ Y : \hat \R $, we need to show that the terminal projection $ ! $ is in $ D(Y, I) $. Note that any morphism trivially commutes with the terminal projections. Since $ Y : \hat \R $, we have $ (r, s) : Y \triangleleft L $, and since $ L \cong I \times L $, this means that the terminal projection is indeed in $ D(Y, I) $:
    \begin{center}
      \begin{tikzcd}
        Y \arrow[rd, "!"'] \arrow[r, shift left, "s"] & L \arrow[d, "!"] \arrow[l, shift left, "r"] \arrow[r, "\sim"] & I \times L \arrow[ld, "!"]\\
        & I
      \end{tikzcd}
    \end{center}
\end{enumerate}

@@

Now, Hyland shows the following:
\begin{theorem}\label{thm:restrict-sum-product}
  For $ f : D(Y, X) $, we can restrict $ \sum_f $ and $ \prod_f $ to functors from $ (\hat \R \downarrow_D Y) $ to $ (\hat \R \downarrow_D X) $.
\end{theorem}
\begin{proof}
  Given $ (Z, g) : (\Pshf L \downarrow Y) $, for $ \prod_F $ we only have to show that if we have $ (Z, g) \triangleleft \Delta_Y L $, then we have $ \prod_f(Z, g) \triangleleft \Delta_X L $. Note that since $ X $ is in $ \hat \R $, this also shows that $ \prod_f Z $ is in $ \hat \R $. Because functors preserve retractions, it suffices to show that we have $ \prod_f \Delta_Y L \triangleleft \Delta_X L $.

@@

  Therefore, we have the following diagram, showing that $ \prod_f \Delta_Y L $ is a retract of $ \prod_{p_1} \Delta_{X \times L} L $:
  \begin{center}
    \begin{tikzcd}
      & \prod_{p_1} \Delta_{X \times L} L\\
      \prod_f \Delta_Y L
        \arrow[r, "\prod_f \eta_r"]
        \arrow[rr, bend right, "\prod_f \eta_{s \cdot r}"]
        \arrow[rrr, bend right, "\id {\Pshf L \downarrow X}"] &
      \prod_f \prod_r r^* \Delta_Y L \arrow[r, "\prod_f \prod_r \eta_s"] \arrow[u, "\sim"] &
      \prod_f \prod_r \prod_s s^* r^* \Delta_Y L \arrow[r, "\sim"] &
      \prod_f \Delta_Y L
    \end{tikzcd}
  \end{center}
  So it suffices to show that $ \prod_{p_1} \Delta_{X \times L} L $ is a retract of $ \Delta_X L $. By Lemma \ref{lem:constant-dependent-product}, we have $ \prod_{p_1} \Delta_{X \times L} L \cong \Delta_X L^L $. By functoriality of $ \Delta_X $, it suffices to show that $ L^L $ is a retract of $ L $. But we already saw in Theorem \ref{thm:representation-theorem} that $ L^L \cong A(L, 1) $ is a retract of $ L $. This shows that $ \prod_f $ indeed restricts to a functor from $ (\hat \R \downarrow_D Y) $ to $ (\hat \R \downarrow_D X) $.

  The proof for $ \sum_f $ is very similar. $ \sum_f \Delta_Y L $ is a retract of $ \sum_{p_1} \Delta_{X \times L} L $:
  \begin{center}
    \begin{tikzcd}
      & & \sum_{p_1} \Delta_{X \times L} L \\
      \sum_f \Delta_Y L
        \arrow[r, "\sim"]
        \arrow[rrr, bend right, "\id {\Pshf L \downarrow X}"] &
      \sum_f \sum_r \sum_s s^* r^* \Delta_Y L
        \arrow[r, "\sum_f \sum_r \epsilon_s"]
        \arrow[rr, bend right, "\sum_f \epsilon_{s \cdot r}"] &
      \sum_f \sum_r r^* \Delta_Y L
        \arrow[u, "\sim"]
        \arrow[r, "\sum_f \epsilon_r"] &
      \sum_f \Delta_Y L
    \end{tikzcd}
  \end{center}
  By Lemma \ref{lem:sum-postcomposition}, the functor $ \sum_{p_1} $ is given by postcomposition, so $ \sum_{p_1} \Delta_{X \times L} L \cong \Delta_X (L \times L) $. As we saw earlier, we have $ L \triangleleft L \times L $, which completes the proof that $ \sum_f $ restricts to a functor from $ (\hat \R \downarrow_D Y) $ to $ (\hat \R \downarrow_D X) $.
\end{proof}

\begin{corollary}\label{cor:relatively-cartesian-closed}
  Since the $ (\hat \R \downarrow_D X) $ are full subcategories of the $ (\Pshf L \downarrow X) $, so we can restrict the adjunction $ f^* \dashv \prod_f $ to an adjunction $ \bar f^* \dashv \bar \prod_f $ between $ (\hat \R \downarrow_D X) $ and $ (\hat \R \downarrow_D Y) $, which shows that $ \hat \R $ is cartesian closed relative to the class $ D $ of display maps.
\end{corollary}

@@

  \enquote{$ F $ preserves $ \mathbf 1_n $, which determines the function space as a retract of the universal. So $ \bar F $ preserves the retract and the result follows.}

>>>>

Now, Scott's representation theorem, which asks whether we can represent any $ \lambda $-theory as the endomorphism theory of some reflexive object in some cartesian closed category, can be answered using the presheaf category.
\begin{theorem}[\coqident{AlgebraicTheories.RepresentationTheorem}{representation_theorem_iso}]\label{thm:representation-theorem}
  Any $ \lambda $-theory $ L $ is isomorphic to the endomorphism $ \lambda $-theory $ E_{\Pshf L}(L) $, with $ L $ viewed as a presheaf.
\end{theorem}
\begin{proof}
  First of all, remember that $ L $ is indeed exponentiable and that $ L^L = A(L, 1) $ (Lemma \ref{lem:plus-p-presheaf}).
  Now, since $ L $ is a $ \lambda $-theory, we have sequences of functions back and forth $ \lambda_n: A(L, 1)_n \to L_n $ and $ \rho_n: L_n \to A(L, 1)_n $. These commute with the $ L $-actions, so they constitute presheaf morphisms and $ E(L) $ is indeed a $ \lambda $-theory.

  Remark \ref{rem:presheaf-Yoneda} gives a sequence of bijections $ \varphi_n : \Pshf L(L^n, L) \cong L_n $ for all $ n $, sending $ f: \Pshf L(L^n, L) $ to $ f(x_1, \dots, x_n) $, and conversely sending $ s: L_n $ to $ ((t_1, \dots, t_n) \mapsto s \bullet (t_1, \dots, t_n)) $. It considers $ \lambda $-terms in $ n $ variables as $ n $-ary functions on the $ \lambda $-calculus. Therefore, it should come as no surprise that $ \varphi $ preserves the $ x_i $, $ \bullet $, $ \rho $ and $ \lambda $, which makes it into an isomorphism of $ \lambda $-theories and this concludes the proof.
\end{proof}
So Hyland shows that the representation theorem follows largely from the fact that the functions from $ L_n $ to itself can be represented by $ L_{n + 1} $, together with the Yoneda lemma for the Lawvere theory associated to $ L $ (Remark \ref{rem:presheaf-Yoneda}).

@@

\begin{remark}
  Actually, in this thesis we stumbled upon another category with a reflexive object that exhibits $ L $ as its endomorphism theory, provided that $ L $ has $ \beta $-equality: In Appendix \ref{ch:weak-cartesian-closed-categories}, we see that the Lawvere theory $ \L $ associated to $ L $ is a weak cartesian closed category, in which $ 1^1 = 1 $. It is then easy to see that the endomorphism theory that we obtain from this reflexive object $ 1 $ is exactly equal to $ L $.

  Note, however, that this is technically not an answer to Scott's original question, because the original formulation asks for a cartesian closed category with a reflexive object, whereas here we only have a reflexive object in a weak cartesian closed category. Even so, it is an answer to the core of the question, which is about whether we can represent any $ \lambda $-theory as the sets of morphisms $ \C(U^n, U) $ for some $ U : \C $.
\end{remark}

@@

Note that if we consider $ L_1 $ as a monoid, with operation $ \bullet $ and unit $ x_1 $, and if we consider this monoid as a category, we can embed this (fully faithfully) into $ \mathbf L $. This gives the following sequence of embeddings:
\begin{center}
  \begin{tikzcd}
    \C_{L_1} \arrow[r, hookrightarrow] & \mathbf L \arrow[r, hookrightarrow] & \R \arrow[r, hookrightarrow] & \Pshf L
  \end{tikzcd}
\end{center}

Note that the composition of the first two morphisms sends the object of the category $ \C_{L_1} $ to $ (\lambda x_1, x_1) : \R $, which is exactly the usual embedding of $ \C_{L_1} $ into its Karoubi envelope.

Write $ i: \op{\C_{L_1}} \hookrightarrow \op{\mathbf L} $ and $ j: \op{\mathbf L} \hookrightarrow \op \R $ for the embeddings on the opposite categories. By Corollary \ref{cor:surjective-precomposition}, the first two functors in this sequence essentially yield surjective functors on the presheaf categories:
\begin{center}
  \begin{tikzcd}
    \C_{L_1} \arrow[d, "\yo", hook] &
    \mathbf L \arrow[l, "\op i"'] \arrow[d, "yo", hook] &
    \mathbf R \arrow[rd, hook] \arrow[d, "\yo", hook] \arrow[l, "\op j"']\\
    P \C_{L_1} &
    P \mathbf L \arrow[l, "i_*"', two heads] &
    P \R \arrow[l, "j_*"', two heads] \arrow[ll, "\sim"', bend left] &
    \Pshf L \arrow[ll, "\sim", bend left]
  \end{tikzcd}
\end{center}
The two equivalences in this diagram are due to Lemma \ref{lem:lawvere-presheaf} and Corollary \ref{cor:karoubi-presheaf}.

@@

Recall that in the last chapter, we saw that Paul Taylor shows that the category of retracts $ \R $, studied by Scott, is not only cartesian closed, but also `relatively cartesian closed'. The proof uses the axiom of choice, and works for the setcategory version of the Karoubi envelope $ \overline{\C_{L_0}} $. Here, $ \C_{L_0} $ is the category associated to the monoid $ (L_0, \circ) $.

Now, recall from Chapter \ref{ch:category-theory}, that in univalent foundations there are three different definitions that we can take for the Karoubi envelope, denoted $ \overline \C_{L_0} $, $ \widetilde \C_{L_0} $ and $ \widehat \C_{L_0} $. There is three of them because in classical category theory there are two definitions that are equivalent, and because for one of the definitions we have to choose whether we interpret the `existence' of a retraction as additional structure (the retraction is given explicitly) or as a property (where we demand `mere existence'). This gives the following diagram (Corollary \ref{cor:karoubi-candidates}):
\begin{center}
  \begin{tikzcd}
    \C_{L_0} \arrow[r, hook, "\iota_C"] & \overline \C_{L_0} \arrow[r, "\sim"] & \widetilde \C_{L_0} \arrow[r, hook, two heads] & \widehat \C_{L_0} \arrow[r, phantom, "\subseteq"] & P \C
  \end{tikzcd}
\end{center}
where $ \overline \C_{L_0} $ and $ \widetilde \C_{L_0} $ are equivalent to each other, but usually not to $ \widehat \C_{L_0} $. We saw that $ \widehat \C_{L_0} $ is the Rezk completion of $ \overline \C_{L_0} $ and $ \widetilde \C_{L_0} $.

@@

The proofs of Scott and Taylor are about $ \overline \C_{L_0} $, because they work explicitly with idempotents. The question whether it is possible to lift the proofs to $ \widetilde \C_{L_0} $ or $ \widehat \C_{L_0} $, is left for future research. On the other hand, as we will see, Hyland works with presheaves and in this way, he shows that $ \widehat \C_{L_0} $ is relatively cartesian closed.

Recall that in this thesis, we use the notation $ \Delta_X $ for the functor $ \C \to (\C \downarrow X) $, sending $ Y $ to $ p_1 : X \times Y \to X $. Also, we use the notation $ X \triangleleft Y $ to denote the type of retraction-section pairs $ r: Y \to X $ and $ s: X \to Y $. We will sometimes not name $ (r, s) : X \triangleleft Y $ explicitly, but just mention that we have $ X \triangleleft Y $, meaning that we have such $ r $ and $ s $.

Now, officially, $ \widehat \C_{L_0} $ is the full subcategory of $ P \C_{L_0} $, consisting of all objects $ X : P \C_{L_0} $ such that there merely exists $ X \triangleleft \yo(\star) $. However, recall from the previous section that $ P \C_{L_0} \simeq P \C_{L_1} \simeq \Pshf L $. Under this equivalence, $ \yo(\star) $ is embedded as the theory presheaf $ L $, and we will work in the full subcategory $ \widehat \R \subseteq \Pshf L $ of the objects $ X : \Pshf L $ such that there merely exists $ X \triangleleft L $. Of course, then $ \widehat \R $ is equivalent to $ \widehat \C_{L_0} $.

@@

  Also, functors preserve retractions, so if we have a functor $ F : \C \to \D $ and we have $ (r, s) : X \triangleleft Y $, then we have
  \[ (F(r), F(s)) : F(X) \triangleleft F(Y). \]

  Lastly, if we want to show that we can do some construction on some $ X : \widehat \R $, we can often borrow the construction from $ \Pshf L $ to get some object $ Y : \Pshf L $ and for this part, we cannot use the $ (r, s) : X \triangleleft L $. Then, to show that $ Y $ is indeed in $ \widehat \R $, we must show that there merely exist $ (r^\prime, s^\prime) : Y \triangleleft L $. By the recursion principle of the propositional truncation, we can assume that we have a concrete $ (r, s) : X \triangleleft L $ to construct $ r^\prime $ and $ s^\prime $.
\end{remark}

Recall that Paul Taylor's display maps for $ \R $ can be characterized as the retracts of $ \Delta_X U $ in $ (\R \downarrow X) $. This idea also works for $ \widehat \R $, and Hyland defines $ D(Y, X) \subseteq \widehat \R(Y, X) $ to consist of the retracts of $ \Delta_X L $ in $ (\widehat \R \downarrow X) $. Note that, because $ \widehat \R $ is a full subcategory of $ \Pshf L $, $ (\widehat \R \downarrow X) $ is equivalent to $ (\Pshf L \downarrow X) $ for $ X : \widehat \R $.

@@

  Therefore, if we want $ D(Y, X) $ to be a subset of $ \widehat \R(Y, X) $, we need to take the propositional truncation of the existence of $ r $ and $ s $.
\end{remark}

\begin{remark}
  Recall that for all $ X : \R $, we have $ X \triangleleft U $, where $ U = \lambda x_1, x_1 $, so by the above, for all $ Y $ in the image of the embedding of $ \R $ into $ \Pshf L $, we have $ Y \triangleleft L $. In particular, $ Y $ is in $ \widehat \R $. Actually, we already knew this, because this essentially describes the embedding of $ \R $ into its Rezk completion $ \widehat \R $.

  An important example is $ U \times U : \R $, which maps to something isomorphic to $ L \times L $ (because the Yoneda embedding and any adjoint equivalence preserves limits). The retraction $ (r, s) : L \times L \triangleleft L $ is given explicitly by the (pairing and splitting) morphisms
  \[ s_n(a, b) = \lambda x_{n + 1}, x_{n + 1} a b \quad \text{and} \quad r_n(a) = (a (\lambda x_{n + 1} x_{n + 2}, x_{n + 1}), a (\lambda x_{n + 1} x_{n + 2}, x_{n + 2})). \]
  Now, for any $ X, Y : \Pshf L $ and $ f : \Pshf L(Y, X) $, if we have $ (Y, f) \triangleleft \Delta_X L $ in $ \Pshf L $, we have in particular $ Y \triangleleft X \times L $. If we also have $ X \triangleleft L $, we also have $ X \times L \triangleleft L \times L $ because $ - \times L $ is a functor. Then we have
  \[ Y \triangleleft X \times L \triangleleft L \times L \triangleleft L. \]
  In particular, for any $ X : \widehat \R $, if some $ (Y, f) : (\Pshf L \downarrow X) $ is a retract of $ \Delta_X L $, we know that $ Y $ is in $ \widehat \R $.
\end{remark}

Now we can show that $ D $ is indeed a class of display maps:
\begin{enumerate}
  \item Take $ f : D(Y, X) $ and $ g: \widehat \R(Z, X) $. We can borrow the pullback from $ \Pshf L $ to get $ g^*(Y, f) : (\Pshf L \downarrow Z) $. We now must show that there merely exists $ g^* (Y, f) \triangleleft \Delta_Z L $ and in this part, we can assume that the mere existences of retractions are actually given by concrete retraction-section pairs.

    Now, note that taking pullbacks gives a functor $ g^* : (\Pshf L \downarrow X) \to (\Pshf L \downarrow Z) $. Now, $ (Y, f) \triangleleft \Delta_X L $, gives $ (g^*Y, g^*f) \triangleleft g^*(\Delta_X L) $:
    \begin{center}
      \begin{tikzcd}
        g^* Y
          \arrow[rd, "g^*s", shift left]
          \arrow[dd, "g^* f"]
          \arrow[rr] &&
        Y
          \arrow[rd, "s", shift left]
          \arrow[dd, "f"] & \\
        & g^* (X \times L)
          \arrow[lu, "g^*r", shift left]
          \arrow[ld, "g^* p_1"]
          \arrow[rr] & &
        X \times L
          \arrow[ld, "p_1"]
          \arrow[lu, "r", shift left] \\
        Z
          \arrow[rr, "g"] & &
        X &
      \end{tikzcd}
    \end{center}
    The isomorphism between $ g^*(\Delta_X L) $ and $ \Delta_Z L $ (Remark \ref{rem:pullback-of-projection}) shows that we have $ (g^*Y, g^*f) \triangleleft \Delta_Z L $. From this, it also follows that $ g^* Y $ is in $ \widehat \R $.

    Now, $ g^*Y $ is the pullback of $ f $ and $ g $ in $ \Pshf L $, and $ \widehat \R $ is a full subcategory of $ \Pshf L $, so $ g^* Y $ is also the pullback of $ f $ and $ g $ in $ \widehat \R $.

@@

  \item Note that $ \Pshf L $ has a terminal presheaf $ I_n = \{ \star \} $. Under the embedding of $ \R $ into $ \Pshf L $, this is (isomorphic to) the image of $ \lambda x_1 x_2, x_2 : \R $, and therefore, it is in $ \widehat \R $. Now, given $ Y : \widehat \R $, we need to show that the terminal projection $ ! $ is in $ D(Y, I) $. Note that any morphism trivially commutes with the terminal projections. Since $ Y : \widehat \R $, we have $ (r, s) : Y \triangleleft L $, and since $ L \cong I \times L $, this means that the terminal projection is indeed in $ D(Y, I) $:
    \begin{center}
      \begin{tikzcd}
        Y \arrow[rd, "!"'] \arrow[r, shift left, "s"] & L \arrow[d, "!"] \arrow[l, shift left, "r"] \arrow[r, "\sim"] & I \times L \arrow[ld, "!"]\\
        & I
      \end{tikzcd}
    \end{center}
\end{enumerate}

@@

Now, Hyland shows the following:
\begin{theorem}\label{thm:restrict-sum-product}
  For $ f : D(Y, X) $, we can restrict $ \sum_f $ and $ \prod_f $ to functors from $ (\widehat \R \downarrow_D Y) $ to $ (\widehat \R \downarrow_D X) $.
\end{theorem}
\begin{proof}
  Given $ (Z, g) : (\Pshf L \downarrow Y) $, for $ \prod_F $ we only have to show that if we have $ (Z, g) \triangleleft \Delta_Y L $, then we have $ \prod_f(Z, g) \triangleleft \Delta_X L $. Note that since $ X $ is in $ \widehat \R $, this also shows that $ \prod_f Z $ is in $ \widehat \R $. Because functors preserve retractions, it suffices to show that we have $ \prod_f \Delta_Y L \triangleleft \Delta_X L $.

@@

  Therefore, we have the following diagram, showing that $ \prod_f \Delta_Y L $ is a retract of $ \prod_{p_1} \Delta_{X \times L} L $:
  \begin{center}
    \begin{tikzcd}
      & \prod_{p_1} \Delta_{X \times L} L\\
      \prod_f \Delta_Y L
        \arrow[r, "\prod_f \eta_r"]
        \arrow[rr, bend right, "\prod_f \eta_{s \cdot r}"]
        \arrow[rrr, bend right, "\id {\Pshf L \downarrow X}"] &
      \prod_f \prod_r r^* \Delta_Y L \arrow[r, "\prod_f \prod_r \eta_s"] \arrow[u, "\sim"] &
      \prod_f \prod_r \prod_s s^* r^* \Delta_Y L \arrow[r, "\sim"] &
      \prod_f \Delta_Y L
    \end{tikzcd}
  \end{center}
  So it suffices to show that $ \prod_{p_1} \Delta_{X \times L} L $ is a retract of $ \Delta_X L $. By Lemma \ref{lem:constant-dependent-product}, we have $ \prod_{p_1} \Delta_{X \times L} L \cong \Delta_X L^L $. By functoriality of $ \Delta_X $, it suffices to show that $ L^L $ is a retract of $ L $. But we already saw in Theorem \ref{thm:representation-theorem} that $ L^L \cong A(L, 1) $ is a retract of $ L $. This shows that $ \prod_f $ indeed restricts to a functor from $ (\widehat \R \downarrow_D Y) $ to $ (\widehat \R \downarrow_D X) $.

  The proof for $ \sum_f $ is very similar. $ \sum_f \Delta_Y L $ is a retract of $ \sum_{p_1} \Delta_{X \times L} L $:
  \begin{center}
    \begin{tikzcd}
      & & \sum_{p_1} \Delta_{X \times L} L \\
      \sum_f \Delta_Y L
        \arrow[r, "\sim"]
        \arrow[rrr, bend right, "\id {\Pshf L \downarrow X}"] &
      \sum_f \sum_r \sum_s s^* r^* \Delta_Y L
        \arrow[r, "\sum_f \sum_r \epsilon_s"]
        \arrow[rr, bend right, "\sum_f \epsilon_{s \cdot r}"] &
      \sum_f \sum_r r^* \Delta_Y L
        \arrow[u, "\sim"]
        \arrow[r, "\sum_f \epsilon_r"] &
      \sum_f \Delta_Y L
    \end{tikzcd}
  \end{center}
  By Lemma \ref{lem:sum-postcomposition}, the functor $ \sum_{p_1} $ is given by postcomposition, so $ \sum_{p_1} \Delta_{X \times L} L \cong \Delta_X (L \times L) $. As we saw earlier, we have $ L \triangleleft L \times L $, which completes the proof that $ \sum_f $ restricts to a functor from $ (\widehat \R \downarrow_D Y) $ to $ (\widehat \R \downarrow_D X) $.
\end{proof}

\begin{corollary}\label{cor:relatively-cartesian-closed}
  Since the $ (\widehat \R \downarrow_D X) $ are full subcategories of the $ (\Pshf L \downarrow X) $, so we can restrict the adjunction $ f^* \dashv \prod_f $ to an adjunction $ \overline f^* \dashv \overline \prod_f $ between $ (\widehat \R \downarrow_D X) $ and $ (\widehat \R \downarrow_D Y) $, which shows that $ \widehat \R $ is cartesian closed relative to the class $ D $ of display maps.
\end{corollary}

@@

  \enquote{$ F $ preserves $ \mathbf 1_n $, which determines the function space as a retract of the universal. So $ \overline F $ preserves the retract and the result follows.}
